\section{Discussion}
\label{sec:discussion}

\subsection{What the multi-instrument representation actually buys}
% --- the honest core ---
The cross-instrument gain is real but bounded. The joint position improves
faint-band localization because it imports the sharpest available
(\textit{Euclid} VIS) constraint; it adds no precision beyond that band, and
where another band is intrinsically sharper, transferring the VIS position is
counterproductive (\S\ref{sec:results:injection}). The practical value is
therefore largest exactly where it is most needed: measuring positions in bands
where a source is too faint to localize on its own.

\subsection{Foundation-version insensitivity}
% --- negative result, stated plainly ---
We trained several foundation variants (\S\ref{sec:methods:foundation}).
In-distribution linear probes and downstream detection/astrometry metrics do
not statistically distinguish them. \todo{Quantify.} We argue the
discriminating test for foundation quality is out-of-distribution transfer
(\S\ref{sec:ood}) rather than in-distribution downstream accuracy, which
saturates against label noise.

\subsection{Label noise and the limits of in-distribution metrics}
\todo{Summarize the centroid-emulation finding: at the faint end the
band-vs-canonical residual measures the model reproducing the canonical
centroid definition, not positional truth; this is why injection and
\textit{Gaia} (truth metrics) are necessary and in-distribution residuals are
not sufficient.}

\subsection{Bright-star detection suppression}
% --- mechanism + fix ---
The signal-to-noise input normalization that stabilizes training also lets a
saturated star dominate the local feature statistics, suppressing faint
detections nearby (\S\ref{sec:results:detection}). This is a completeness, not
an astrometric-accuracy, effect. Mitigations --- local background subtraction,
a bright-star-aware rescue pass, or a normalization less sensitive to saturated
outliers --- are left to future work. \todo{If MER shows the lost sources are
real, quantify the completeness cost and any mitigation tested.}

\subsection{Limitations}
\begin{itemize}
  \item Injection uses idealized PSFs and isolated sources; blends and
    realistic wings are not captured.
  \item Out-of-distribution evidence is from a single second field.
  \item \todo{variance-map availability; Rubin optical transfer; others.}
\end{itemize}
