\section{Introduction}
\label{sec:intro}

% --- the opportunity ---
Rubin and \textit{Euclid} image overlapping sky with complementary strengths:
Rubin provides deep six-band optical photometry at ground-based resolution,
while \textit{Euclid} VIS delivers diffraction-limited optical morphology and
NISP adds near-infrared coverage. Jointly they constrain spectral energy
distributions and positions better than either alone, yet standard pipelines
process each instrument independently and combine only at the catalog level,
discarding pixel-level cross-instrument information.

% --- foundation models in astronomy: position relative to prior work ---
\todo{Position relative to prior astronomical foundation / multimodal
representation models (e.g.\ AstroCLIP, Multimodal Universe, AstroPT) and to
classical multi-instrument forced photometry / forced astrometry. Make the
novelty boundary explicit: pixel-level, multi-instrument, mixed-resolution
fusion driving downstream detection and astrometry.}

% --- what this paper does ---
In this paper we (i) describe \jaisp{}, a joint Rubin--\textit{Euclid}
foundation model trained by masked reconstruction on a common fused pixel grid
(\S\ref{sec:methods}); (ii) use the frozen representation for source detection
and for a multi-band joint-centroid astrometry that transfers the sharp
\textit{Euclid} VIS localization to bands where the source is faint
(\S\ref{sec:results}); (iii) validate positions against truth via
source injection and against \textit{Gaia}; and (iv) report a first
out-of-distribution evaluation on a second field with no retraining
(\S\ref{sec:ood}). Throughout we take a deliberately critical stance, using
ablations and controlled injections to separate what the multi-instrument
representation genuinely buys from what is already available in the single
sharpest band.

% --- contributions ---
Our main contributions are:
\begin{enumerate}
  \item A multi-instrument, mixed-resolution pixel-level foundation model for
    Rubin$+$\textit{Euclid}, with frozen-feature downstream heads.
  \item A multi-band joint-centroid astrometry method and a controlled
    quantification, via source injection, of the cross-instrument localization
    gain and its VIS-imposed ceiling.
  \item A first transfer (out-of-distribution) evaluation of the frozen stack
    on an independent field.
  \item Honest characterization of failure modes: foundation-version
    insensitivity, the VIS localization floor, and bright-star detection
    suppression.
\end{enumerate}
