\section{Methods}
\label{sec:methods}

\subsection{The \jaisp{} foundation model}
\label{sec:methods:foundation}

% --- design path: what we tried and abandoned ---
The current design follows from a sequence of approaches we tried and
discarded, and the path is itself one of the paper's lessons. We first pursued
\emph{latent-space alignment} objectives that pull co-located Rubin and
\textit{Euclid} features together: patch-level contrastive learning, an
object-centric DETR-style variant with learned object queries, and
JEPA-style student--teacher matching at the token level, including a
strict-position variant. Each failed for precision work, for a common reason.
Contrastive patch embeddings collapsed onto the sky background that dominates
$\sim$95\% of the pixels; the object-centric matching was unstable in crowded
deep fields; and the token-level objectives, even with exact positional
matching, rewarded feature \emph{similarity} without requiring the encoder to
preserve where flux sits to sub-pixel accuracy --- the coarse ($\sim$3\arcsec)
patch tokenization compounded this. The decisive evidence was that a simple
supervised cost-volume CNN reached $\sim$38$\mas$ on the astrometry task while
the best JEPA features gave only $\sim$47$\mas$: the expensive self-supervised
representation was worse than a small supervised baseline. We concluded that
latent-similarity objectives do not retain the spatial information precision
cosmology needs, and moved to \emph{pixel-space masked reconstruction}, where
holding out a band and reconstructing its pixels forces the encoder to encode
exact positions. A fuller account of the abandoned variants is given in the
project documentation.

% --- architecture ---
\jaisp{} encodes each band with a per-band convolutional stem, resamples the
per-instrument streams to a common fused pixel scale of $0.4\arcsecpx$, and
fuses them into a shared bottleneck representation. \todo{Architecture detail:
per-band stems, mixed-resolution stream depths from $\log_2$(scale ratio),
ConvNeXt/transformer blocks, FiLM band conditioning, concat fusion (v9/v10).}
The per-band stem normalizes its input by the per-pixel noise,
$x = \mathrm{image}/\mathrm{rms}$, and clamps the result to a fixed
signal-to-noise range before feature extraction; this choice is revisited in
\S\ref{sec:results} as the origin of bright-star detection suppression.

% --- training ---
The model is trained by masked reconstruction \todo{(MAE objective, masking
fraction, cross-instrument masking, Charbonnier$+$core-L2 loss for v10), epochs,
optimizer, hardware}. All downstream heads operate on \emph{frozen} features.

% --- versions / equivalence note ---
Within the reconstruction era the architecture was refined in several steps:
an initial dense variant downsampled \textit{Euclid} VIS to the Rubin grid and
discarded its resolution advantage; this was fixed by a mixed-resolution
two-stream encoder that keeps each instrument at native scale, followed by a
finer fused scale ($0.4\arcsecpx$ via cropped inputs), symmetric per-band
concatenation fusion, and a Charbonnier reconstruction loss with extra
weighting on high-signal core pixels. We report the production checkpoint used
throughout (the loss-shaped, concat-fusion variant). In-distribution probes and
downstream detection/astrometry metrics do not statistically distinguish the
last few variants (\S\ref{sec:discussion}); we therefore do not claim a winner
among them on this field and treat out-of-distribution transfer as the
discriminating test. \todo{Fix final version nomenclature/checkpoint name and
training config.}

\subsection{Detection head}
\label{sec:methods:detection}

% --- centernet ---
We detect sources with a center-prediction (CenterNet-style) head on the frozen
fused features, producing a normalized centroid and score per source.
\todo{Head architecture, self-training/pseudo-label procedure, operating point
(conf threshold), NMS.} Detections are gated on the \textit{Euclid} VIS signal
and serve as seeds for astrometry.

\subsection{Multi-band joint-centroid astrometry}
\label{sec:methods:astrometry}

% --- the instrument ---
Given a detection seed, we (i) refine the centroid classically on
\textit{Euclid} VIS, (ii) solve for a single \emph{joint canonical} sky
position by simultaneously fitting the source across all bands, weighting each
band by its noise and accounting for its PSF, and (iii) measure each band's
classical centroid and record its offset from the joint canonical position.
The joint canonical position is the cross-instrument estimate; the per-band
offset is the astrometric residual we report.

% --- why this beats single-band ---
Because the fit is dominated by the band with the sharpest, highest-SNR
localization (typically \textit{Euclid} VIS), the joint position transfers that
localization to bands in which the source is faint or unresolved
(\S\ref{sec:results}). Per-source uncertainties follow from the $\chi^2$
curvature of the joint fit and a per-band $\mathrm{FWHM}/(2.355\,\mathrm{SNR})$
centroid-noise model.

% --- foundation-agnostic note ---
We emphasize that step (i)--(iii) are classical given the detection seeds; the
foundation's role in astrometry is to supply the cross-instrument detections.
A learned per-object position head was also trained and is discussed as an
ablation in \S\ref{sec:discussion}.
